Computing volume via non-uniform rectangular infinitesimals

### 2019-08-16 Cav2D/article-template.tex
fixed non-linear reference

\noindent
Figure~\ref{fig:cav_principle} depicts a classic rectangular prism or column (left solid) as well as a \emph{non-uniform rectangular prism or column} (right solid). The volumes of the two solids in Figure~\ref{fig:cav_principle} are equal due 
to Cavalieri's principle. Note that in this article we use the non-standardized phrase non-unifor rectangular prism or column to refer to a solid that has rectangular 
basis and lateral edges that are either oblique lines or non-linear curves. Moreover, the four lateral edges of the aformentioned solid also need to be translations of one another. In this paper, we will also take the liberty to refer to a non-linear rectangular prism whose base's
lenght and widht both tend to zero as a \emph{non-uniform rectangular infinitesimal}. \\

\noindent
Ackermann et al. utilized the idea of using non-rectangular integration strips to perform integration in $\mathbb{R}^2$ and in doing so developed the Cavalieri integral \cite{ackermann12}. 
The area of a non-rectangular integration strip is computed via Cavalieri’s principle. The Cavalieri integral has proven to be very useful: it can be used to visualize the Riemann-Stieltjes integral \cite{grobler19}. In this paper, we extend the Cavalieri integral into $\mathbb{R}^3$, i.e. we use Cavalieri's principle in $\mathbb{R}^3$ to define the double 
Cavalieri integral. We also show that if a solid can be decomposed into an infinite number of non-uniform rectangular infinitesimals then its volume can be computed 
by the double Cavalieri integral. Whether the double Cavalieri integral will prove to be an equally useful visualization tool will be investigated as part of a future endeavor.
Moreover, the examples we provide at the end of the paper are presented in such a way that they will deepen undergraduate Calculus students' understanding into the 
process of integration. As this paper is mainly meant to be read by undergraduate students as well as the layman who is interested in the theory of integration we will 
try to steer clear from using too many formal definitions and proofs and rather convey our ideas via worked out examples.\\

@@

\section{Double Cavalieri Integral}
In this section we define the double Cavalieri integral and we show that it can be used to calculate the volume of solids that can be decomposed into an infinite 
number of non-linear regtangular infinitesimals.\\

@@

%We will now show 
%that we can approximate $V_{\mathcal{S}}$ by summing together the volume of $n$ non-linear prisms that cicumscribe $\mathcal{S}$ and that as $n\rightarrow \infty$ 
%the aformentioned sum approaches $V_{\mathcal{S}}$. 
%Without any loss of generality, assume that $f(x,y)\geq0$. Also, let $S_f = \{(x,y,z)\in\mathbb{R}^3|(x,y)\in S ~ \textrm{and} ~ z=f(x,y)\}$. 
%Let $\mathcal{S}$ denote the solid which is obtained by connecting the vertices of $R$ and $S_f$ via translations of $\mathbf{c}(t)$ along 
%the $xy$ plane.

@@

\noindent
Also, let $S_f = \{(x,y,z)\in\mathbb{R}^3|(x,y)\in S ~ \textrm{and} ~ z=f(x,y)\}$.  Let us further assume that $f(x,y) > 0$ for all $(x,y)\in S$. Let $\mathcal{S}$ denote the solid which is obtained by connecting the vertices of $R$ and 
$S_f$ via translations of $\mathbf{c}(t)$ along the $xy$ plane and then closing the resulting open faces that form due to the aforementioned connections by employing appropriate surfaces. Note that 
$\mathcal{S}$ can be decomposed into an infinite number of non-uniform rectangular infinitesimals. As an example, the solid $\mathcal{S}$ depicted in Figure~\ref{fig:prismatoid_solid} is obtained if we set $f(x,y) = -x - y + 8$, $\mathbf{c}(t) = -\left<\frac{1}{2}t,t,t\right>$ and $R = \{(x,y)\in\mathbb{R}^2|0 \geq x \geq 4, 0 \geq x \geq 4\} \nonumber$.
Moreover, the domain $R$ and the range $S$ of the transformation function $T$ which is associated with the solid $\mathcal{S}$ is depicted in Figure~\ref{fig:prismatoid_regions}.  
In general, let $V_{\mathcal{S}}$ denote the volume of $\mathcal{S}$. We can approximate $V_{\mathcal{S}}$ with:
\begin{equation}
\label{eq:cav_sum_2d}
V_{\mathcal{S}} \approx \sum_{i=0}^{n-1} \sum_{j=0}^{m-1} f(x_i^2,y_j^2) \Delta x_i^1 \Delta y_j^1.
\end{equation}
Let us verify that this is indeed the case. If we closely inspect Equation~\ref{eq:cav_sum_2d} and Figure~\ref{fig:prismatoid_solid} we notice the following:
\begin{itemize}
\item $\Delta x_i^1$ and $\Delta y_j^1$ represent the width and the length of subrectangle $R_{ij}$.
\item The product $\Delta A = \Delta x_i^1\times \Delta y_j^1$ represents the area of subrectangle $R_{ij}$.
\item $f(x_i^2,y_j^2)$ represents the height of a non-linear rectangular column or prism $\mathcal{K}_{ij}$ whose base is $R_{ij}$. Furthermore, the lateral edges of this column are all translations 
of $\mathbf{c}(t)$ along the $xy$-plane. The cross-section between any parallel plane to the $xy$-plane and $\mathcal{K}_{ij}$ is a rectangle with the same dimensions as $R_{ij}$. 
\item The product $f(x_i^2,y_j^2)\times \Delta A$ represents the volume of $\mathcal{K}_{ij}$ (due to Cavalieri's principle).
\item The double Cavalieri sum $\sum_{i=0}^{n-1} \sum_{j=0}^{m-1} f(x_i^2,y_j^2) \Delta A$, therefore, approximates $V_{\mathcal{S}}$. 
\item In the limit the double Cavalieri sum approaches $V_{\mathcal{S}}$ (if this limit exists). 
\end{itemize}

@@

\noindent
Since $\mathcal{S}$ can be decomposed into an infinite number of non-linear rectangular infinitesimals we can compute its volume using:
\begin{equation}
\label{eq:cav_ex2}
V_{\mathcal{S}} = \iint\limits_{\!\mathcal{C}} f(x,y) dA =  \int_a^b\int_c^d f(h(x,y),s(x,y))~dx~dy,
\end{equation}
with $\mathcal{C}=\{\mathbf{c}(t),R\}$. We, however, first need to calculate $T$ if we wish to evaluate Equation~\ref{eq:cav_ex2} (see Figure~\ref{fig:sphere_regions}). We can determine $T$ via 
Equation~\eqref{eq:h_2d} and Equation~\eqref{eq:s_2d}. Equation~\eqref{eq:h_2d} and Equation~\eqref{eq:s_2d} in turn requires $t(x_i^1,y_j^2)$, which can be obtained via Equation~\eqref{eq:c_z}.
If we substite $f(x,y)$ and $\mathbf{c}(t)$ into Equation~\eqref{eq:c_z} and then solve for $t$ we obtain:
\begin{equation}
t = \frac{1}{4}[-2 x_{i}^1 - 2 y_{j}^1  -  1 + K(x_{i}^1,y_{j}^1)],
\end{equation}
where
\begin{equation}
K(x_{i}^1,y_{j}^1) = \sqrt{-4(x_{i}^1)^2+4x_{i}^1+8x_{i}^1y_{j}^1-4(y_{j}^1)^2+4y_{j}^1 + 33}. 
\end{equation}
As mentioned earlier, we can use $t(x_i^1,y_j^1)$ to find expressions for $h$ and $s$. The expressions of $h$ and $s$ so obtained can then be used to evaluate Equation~\eqref{eq:cav_ex2}:  
\begin{eqnarray}
\iint\limits_{\!\mathcal{C}} f(x,y) dA &=& \int_0^1\int_0^1 f(h(x,y),r(x,y))~dydx\\
&=& \int_0^1 \int_0^1 \sqrt{-(x+t(x,y))^2 - (y+t(x,y))^2 + 4} ~dydx\\
&=& 0.864043
\end{eqnarray}

>>>>

\noindent
Figure~\ref{fig:cav_principle} depicts a classic rectangular prism or column (left solid) as well as a \emph{non-uniform rectangular prism or column} (right solid). The volumes of the two solids in Figure~\ref{fig:cav_principle} are equal due 
to Cavalieri's principle. Note that in this article we use the non-standardized phrase non-unifor rectangular prism or column to refer to a solid that has rectangular 
basis and lateral edges that are either oblique lines or non-linear curves. Moreover, the four lateral edges of the aformentioned solid also need to be translations of one another. In this paper, we will also take the liberty to refer to a non-uniform rectangular prism whose base's
lenght and widht both tend to zero as a \emph{non-uniform rectangular infinitesimal}. \\

\noindent
Ackermann et al. utilized the idea of using non-rectangular integration strips to perform integration in $\mathbb{R}^2$ and in doing so developed the Cavalieri integral \cite{ackermann12}. 
The area of a non-rectangular integration strip is computed via Cavalieri’s principle. The Cavalieri integral has proven to be very useful: it can be used to visualize the Riemann-Stieltjes integral \cite{grobler19}. In this paper, we extend the Cavalieri integral into $\mathbb{R}^3$, i.e. we use Cavalieri's principle in $\mathbb{R}^3$ to define the double 
Cavalieri integral. We also show that if a solid can be decomposed into an infinite number of non-uniform rectangular infinitesimals then its volume can be computed 
by the double Cavalieri integral. Whether the double Cavalieri integral will prove to be an equally useful visualization tool will be investigated as part of a future endeavor.
Moreover, the examples we provide at the end of the paper are presented in such a way that they will deepen undergraduate Calculus students' understanding into the 
process of integration. Furhtermore, we will try to steer clear from using too many formal definitions and proofs and rather convey our ideas via worked out examples, as this paper is mainly meant to be read by undergraduate students as well as the layman who is interested in the theory of integration.\\

@@

\section{Double Cavalieri Integral}
In this section we define the double Cavalieri integral and we show that it can be used to calculate the volume of solids that can be decomposed into an infinite 
number of non-uniform regtangular infinitesimals.\\

@@

%We will now show 
%that we can approximate $V_{\mathcal{S}}$ by summing together the volume of $n$ non-uniform prisms that cicumscribe $\mathcal{S}$ and that as $n\rightarrow \infty$ 
%the aformentioned sum approaches $V_{\mathcal{S}}$. 
%Without any loss of generality, assume that $f(x,y)\geq0$. Also, let $S_f = \{(x,y,z)\in\mathbb{R}^3|(x,y)\in S ~ \textrm{and} ~ z=f(x,y)\}$. 
%Let $\mathcal{S}$ denote the solid which is obtained by connecting the vertices of $R$ and $S_f$ via translations of $\mathbf{c}(t)$ along 
%the $xy$ plane.

@@

\noindent
Also, let $S_f = \{(x,y,z)\in\mathbb{R}^3|(x,y)\in S ~ \textrm{and} ~ z=f(x,y)\}$.  Let us further assume that $f(x,y) > 0$ for all $(x,y)\in S$. Let $\mathcal{S}$ denote the solid which is obtained by connecting the vertices of $R$ and 
$S_f$ via translations of $\mathbf{c}(t)$ along the $xy$ plane and then closing the resulting open faces that form due to the aforementioned connections by employing appropriate surfaces. Note that 
$\mathcal{S}$ can be decomposed into an infinite number of non-uniform rectangular infinitesimals. As an example, the solid $\mathcal{S}$ depicted in Figure~\ref{fig:prismatoid_solid} is obtained if we set $f(x,y) = -x - y + 8$, $\mathbf{c}(t) = -\left<\frac{1}{2}t,t,t\right>$ and $R = \{(x,y)\in\mathbb{R}^2|0 \geq x \geq 4, 0 \geq x \geq 4\} \nonumber$.
Moreover, the domain $R$ and the range $S$ of the transformation function $T$ which is associated with the solid $\mathcal{S}$ is depicted in Figure~\ref{fig:prismatoid_regions}.  
In general, let $V_{\mathcal{S}}$ denote the volume of $\mathcal{S}$. We can approximate $V_{\mathcal{S}}$ with:
\begin{equation}
\label{eq:cav_sum_2d}
V_{\mathcal{S}} \approx \sum_{i=0}^{n-1} \sum_{j=0}^{m-1} f(x_i^2,y_j^2) \Delta x_i^1 \Delta y_j^1.
\end{equation}
Let us verify that this is indeed the case. If we closely inspect Equation~\ref{eq:cav_sum_2d} and Figure~\ref{fig:prismatoid_solid} we notice the following:
\begin{itemize}
\item $\Delta x_i^1$ and $\Delta y_j^1$ represent the width and the length of subrectangle $R_{ij}$.
\item The product $\Delta A = \Delta x_i^1\times \Delta y_j^1$ represents the area of subrectangle $R_{ij}$.
\item $f(x_i^2,y_j^2)$ represents the height of a non-uniform rectangular column or prism $\mathcal{K}_{ij}$ whose base is $R_{ij}$. Furthermore, the lateral edges of this column are all translations 
of $\mathbf{c}(t)$ along the $xy$-plane. The cross-section between any parallel plane to the $xy$-plane and $\mathcal{K}_{ij}$ is a rectangle with the same dimensions as $R_{ij}$. 
\item The product $f(x_i^2,y_j^2)\times \Delta A$ represents the volume of $\mathcal{K}_{ij}$ (due to Cavalieri's principle).
\item The double Cavalieri sum $\sum_{i=0}^{n-1} \sum_{j=0}^{m-1} f(x_i^2,y_j^2) \Delta A$, therefore, approximates $V_{\mathcal{S}}$. 
\item In the limit the double Cavalieri sum approaches $V_{\mathcal{S}}$ (if this limit exists). 
\end{itemize}

@@

\noindent
Since $\mathcal{S}$ can be decomposed into an infinite number of non-uniform rectangular infinitesimals we can compute its volume using:
\begin{equation}
\label{eq:cav_ex2}
V_{\mathcal{S}} = \iint\limits_{\!\mathcal{C}} f(x,y) dA =  \int_a^b\int_c^d f(h(x,y),s(x,y))~dx~dy,
\end{equation}
with $\mathcal{C}=\{\mathbf{c}(t),R\}$. We, however, first need to calculate $T$ if we wish to evaluate Equation~\ref{eq:cav_ex2} (see Figure~\ref{fig:sphere_regions}). We can determine $T$ via 
Equation~\eqref{eq:h_2d} and Equation~\eqref{eq:s_2d}. Equation~\eqref{eq:h_2d} and Equation~\eqref{eq:s_2d} in turn requires $t(x_i^1,y_j^2)$, which can be obtained via Equation~\eqref{eq:c_z}.
If we substite $f(x,y)$ and $\mathbf{c}(t)$ into Equation~\eqref{eq:c_z} and then solve for $t$ we obtain:
\begin{equation}
t = \frac{1}{4}[-2 x_{i}^1 - 2 y_{j}^1  -  1 + K(x_{i}^1,y_{j}^1)],
\end{equation}
where
\begin{equation}
K(x_{i}^1,y_{j}^1) = \sqrt{-4(x_{i}^1)^2+4x_{i}^1+8x_{i}^1y_{j}^1-4(y_{j}^1)^2+4y_{j}^1 + 33}. 
\end{equation}
As mentioned earlier, we can use $t(x_i^1,y_j^1)$ to find expressions for $h$ and $s$. The expressions of $h$ and $s$ so obtained can then be used to evaluate Equation~\eqref{eq:cav_ex2}:  
\begin{eqnarray}
\iint\limits_{\!\mathcal{C}} f(x,y) dA &=& \int_0^1\int_0^1 f(h(x,y),r(x,y))~dydx\\
&=& \int_0^1 \int_0^1 \sqrt{-(x+t(x,y))^2 - (y+t(x,y))^2 + 4} ~dydx\\
&=& 0.864043
\end{eqnarray}
